\chapter{Mathematical Programming}
\localtableofcontents

This document is a result of my pursuits in OR. During my
Bachelor's and Master's studies I have grown fond of operations research, more
specifically linear, integer and stochastic programming. Over the last few years
I have accumulated pile on notes that I thought I would digitalize and share
them with anyone who \st{cares} wants to use them.\\

\section{Language}

These notes were written when I was a student, so if you are looking for a very
dense text on operations research I recommend looking through the references.
I adopt a very informal tone in these notes which I think reflects the fact that
they were written as an inquisitve soul who did not yet have the insight
required to quickly graps very complex defitions and proofs by merely skimming
a couple of lines of text.\\

\section{Background}

%I am a computer scientist, and since I'm writing a text on mathematics I would
%like to emphasize this. I am a \textit{computer scientist}. This may invovle me
%leaving out some of the intricacies of mathematics (e.g. "yes, I STILL
%mean $\forall i,j \in V$..."), but hopefully nothing too serious
%(e.g. "Sure, a partition can be empty!")\footnote{Need proof of this, read up on
%topolofy"}.\\

These notes should help you in understanding operations research
through the eyes of your peer, and hopefully you will think it fun. Bear over
with my tongue-in-cheek writing style and I am sure you can get something out
of it. If not, enjoy the {\LaTeX  } formatting, which is something every
can appreciate.\\

For anyone at the University of Copenhagen with an interest in OR I can 
recommend the following courses:

\begin{description}
\item[\href{http://kurser.ku.dk/course/nmaa05087u/2013-2014}{Operations Research 1}]
\hfill\\ Introduction to OR, mainly linear programming.
\item[\href{http://kurser.ku.dk/course/nmaa09044u/2013-2014}{Operations Research 2}] 
\hfill\\ Integer programming, complexity considerations
and heuristics.
\item[\href{http://kurser.ku.dk/course/nmak13015u/2013-2014}{Modelling and GAMS}
\footnote{I have not taken this course but it has been recommended to me.}]
\hfill\\ (guess what this one is about).
\item[\href{http://kurser.ku.dk/course/nmaa13028u/2013-2014}{Stochastic Programming}]
\hfill\\ (guess what this one is about).
\item[\href{http://kurser.ku.dk/course/ndaa09023u/2013-2014}
{Advanced Algorithms and Data Structures}]
\hfill\\ A great course at DIKU covering
 a plethora of interesting stuff like max flow, \np-completeness, FTP,
 heuristics and randomized algorithms.
\end{description}


{\emph{Non-anticipativity}}{The decision $x_t$, chosen at stage $t$, may depend on the information 
$\xi_{[t]}$, but not on future observations. Equation \ref{eq:nonanti} shows how this is enforced
in the multi-stage case.}

{\emph{Fixed recourse}}{The $\bm{W}$ matrix does not depend on the realization of $\xi$.}

{\emph{Simple recourse}}{Something}

{\emph{Relatively Complete Recourse}}{$K_1 \subset K_2$, so $\mathcal{Q}(x,\xi)$ will never
attain the value $\infty$. In other words, we can always find a feasible second-stage solution
regardless of what choice we make in the first stage.}

{\emph{Complete Recourse}}{$\forall x \in \mathcal{X},\; \exists y \in \mathcal{R}^{n_2} : Wy = z$.
This is a property of $W$, which basically states that if the columns of $W$ span $\mathcal{R}^{n_2}$,
complete recourse applies. It means that we can always find a $y$ regardless of our first-stage
decision. Typically, adding $\big[I, -I\big]$ is sufficienient.}


\input{math_prog/lp}
\input{math_prog/ip}
\input{math_prog/sp}
\input{math_prog/hier}